---
title: "ex3"
author: "Jostein Gjesdal, Jesper Bengston"
date: "2/28/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(ISLR)
library(GGally)
library(ggfortify)
Autosub =  subset(Auto, select =-name)
# Autosub$origin = factor(Autosub$origin)
ggpairs(Autosub)

cor(Autosub)
Auto.lm= lm(mpg ~ . , data= Autosub)
summary(Auto.lm)
autoplot(Auto.lm, smooth.colour = NA)

```
1c)
i: some of the predictors seem to be related to the response. 
ii: spesifically year, origin displacement and weight
iii: the coeficient for year sugests that every year cars get 0.777 higher mpg.

  
1d)
the fitted vs residuals shows a clear pattern tending towards the lower right. This sugests we do not have a true linear dependency in our model. QQ plot looks good. obs 14 seems to have a high leverage. 
```{r}
Autosub2 =  subset(Auto, select =-name)
Autosub2$origin = factor(Autosub2$origin)
ggpairs(Autosub2)


Auto.lm2= lm(mpg ~ displacement + weight + year *origin , data= Autosub2)
summary(Auto.lm2)
anova(Auto.lm2)
```


$$
\hat \beta = \mathbf{(X^T X)}^{-1} X^T Y}

$$

\begin{align}
 E[\hat\beta]=E[(\mathbf{X^T X)^{-1}X^T Y}] \\

\end{align}
